---
title: "R Coding Exercise"
output:
  html_document:
    toc: FALSE
---

*This page is the introductory coding exercise in R for this class. I
will be working with the dslabs package data to perform data processing,
visual plotting, and model fitting.*

## Receiving and summarizing data

```{r Packages}
#| output: false
#Load package
library(dslabs)
library(tidyverse)
library(dplyr)
```

```{r Gapminder Summary}
#Look at help file for gapminder dataset
help(gapminder)
#The gapminder dataset contains health 
#and income outcomes for 184 countries 
#from 1960 to 2016.

#get an overview of the data structure
str(gapminder)

#get a summary of data
summary(gapminder)

#determine the type of object gapminder is
class(gapminder)
```

## Subset data

```{r Making africadata}
#pull out data points (rows) 
#related to Africa and assign to object `africadata`
africadata <- filter(gapminder, continent == "Africa")

#check new object
str(africadata)
summary(africadata)
```

Note that only rows where Africa is listed as the continent are in this
dataset (2907 observations)

```{r Subset columns}
#Now, pull out the columns 
#for infant mortality & life
#expectancy and population & 
#life expectancy for African 
#countries and assign to objects
MortExpec <- africadata %>% select(infant_mortality, life_expectancy)

PopExpec <- africadata %>% select(life_expectancy, population)

#view MortExpec object
str(MortExpec)
summary(MortExpec)

#view PopExpec object
str(PopExpec)
summary(PopExpec)

```

For both new objects, all observations were used (2907 rows) but each
object has two columns (two variables each). All other variables
(including country) have been removed.

## Plotting the data

```{r Scatter plots}
#| label: fig1
#| fig-cap: Life expectancy vs infant mortality
#plot life expectancy as 
#a function of infant mortality
ggplot(MortExpec) +
  geom_point(aes(infant_mortality, life_expectancy))

#| label: fig2
#| fig-cap: Life expectancy vs population size
#plot life expectancy as 
#a function of population size
ggplot(PopExpec) +
  geom_point(aes(population, life_expectancy))+
  scale_x_continuous(trans = "log")

```

Life expectancy has a negative correlation with infant mortality. As
infant mortality increases, life expectancy decreases. The warning
message indicated 226 data points could not be plotted.

Life expectancy has a positive correlation with population size. As
population increases, life expectancy increases. Note that some data
points seem to trend together; this is because the grouped data is from
the same country over multiple years. The warning message indicated that
51 points could not be plotted.

## Data cleaning

As noted in the plotting stage, there are some missing data points for
some countries for some years. Let's try to find the years with missing
data.

```{r Finding NAs}
#find years with missing 
#data for infant mortality 
#using `africadata`

africadata %>%    filter(is.na(infant_mortality))
```

It looks like there is missing data for years 1960 to 1981 and 2016. To
look at a full dataset for one year, we will focus on year 2000.

```{r}
#subset`africadata` to only 
#view year 2000

fullyear <- africadata %>% filter(year == 2000)
str(fullyear)
summary(fullyear)
```

`fullyear` contains only data for year 2000 for the African countries.
This is shown through the `summary` as the statistics all equal 2000 for
the year column. The new dataset had 51 observations with 9 columns.

## Plotting with full dataset (Year 2000)

We will plot the same scatterplots as above with the `fullyear` dataset.
There should not be any missing data, and no warning should pop up from
removing data points.

```{r Scatter plots pt 2}
#| label: fig3
#| fig-cap: Year 2000 life expectancy vs infant mortality 
#plot life expectancy as 
#a function of infant mortality with no missing data
ggplot(fullyear) +
  geom_point(aes(infant_mortality, life_expectancy))

#| label: fig4
#| fig-cap: Year 2000 life expectancy vs population size
#plot life expectancy as 
#a function of population size with no missing data
ggplot(fullyear) +
  geom_point(aes(population, life_expectancy))+
  scale_x_continuous(trans = "log")

```

Based on these figures, we see a similar negative correlation between
life expectancy and infant mortality as in the previous graphs. However,
there does not appear to be a correlation between population size and
life expectancy when looking at year 2000. To use a more objective
measure to observe the relationship between population size and life
expectancy, let's jump into statistical measures.

## Linear modeling

```{r} 
#fitting linear model for population size
#predicting life expectancy
fit1 <- lm(life_expectancy~infant_mortality, data = fullyear)
fit2 <- lm(life_expectancy~population, data = fullyear)

#printing results of the linear models
summary(fit1)
summary(fit2)
```
Looking at the summary results, infant mortality appears to be a valid predictor of life expectancy (p-value<0.001). Life expectancy decreases by 0.189 years with each increase in infant mortality (SE = 0.029, t = -6.594). 

Population size is not a valid predictor of life expectancy because the p-value is 0.616 which is not statistically significant.









