---
title: "R Coding Exercise"
output:
  html_document:
    toc: FALSE
---

*This page is the introductory coding exercise in R for this class. I will be working with the dslabs package data to perform data processing, visual plotting, and model fitting.*

## Receiving and summarizing data

```{r Packages}
#| output: false
#Load package
library(dslabs)
library(tidyverse)
library(dplyr)
```

```{r Gapminder Summary}
#Look at help file for gapminder dataset
help(gapminder)
#the gapminder dataset contains health and income outcomes for 184 countries from 1960 to 2016.

#get an overview of the data structure
str(gapminder)

#get a summary of data
summary(gapminder)

#determine the type of object gapminder is
class(gapminder)
```

## Subset data

```{r Making africadata}
#pull out data points (rows) related to Africa and assign to object `africadata`
africadata <- filter(gapminder, continent == "Africa")

#check new object
str(africadata)
summary(africadata)
```

Note that only rows where Africa is listed as the continent are in this dataset (2907 observations)

```{r Subset columns}
#Now, pull out the columns for infant mortality & life expectancy and population & life expectancy for African countries and assign to objects
MortExpec <- africadata %>% select(infant_mortality, life_expectancy)

PopExpec <- africadata %>% select(life_expectancy, population)

#view MortExpec object
str(MortExpec)
summary(MortExpec)

#view PopExpec object
str(PopExpec)
summary(PopExpec)

```

For both new objects, all observations were used (2907 rows) but each object has two columns (two variables each). All other variables (including country) have been removed.

## Plotting the data

```{r Scatter plots}
#| label: fig1
#| fig-cap: Life expectancy vs infant mortality
#plot life expectancy as a function of infant mortality
ggplot(MortExpec) +
  geom_point(aes(infant_mortality, life_expectancy))

#| label: fig2
#| fig-cap: Life expectancy vs population size
#plot life expectancy as a function of population size
ggplot(PopExpec) +
  geom_point(aes(population, life_expectancy))+
  scale_x_continuous(trans = "log")

```

@fig1 show that life expectancy has a negative correlation with infant mortality. As infant mortality increases, life expectancy decreases. The warning message indicated 226 data points could not be plotted.

@fig2 shows that life expectancy has a positive correlation with population size. As population increases, life expectancy increases. Note that some data points seem to trend together; this is because the grouped data is from the same country over multiple years. The warning message indicated that 51 points could not be plotted.

## Data cleaning

As noted in the plotting stage, there are some missing data points for some countries for some years. Let's try to find the years with missing data.

```{r Finding NAs}
#find years with missing data using `africadata`

```













