---
title: "Exploration"
output: html_document
---

The `Wrangling` script cleaned the raw data from the flu analysis. We will explore the data and look at the symptomatic variables. The main outcomes of interest are `Body Temperature` and `Nausea`. The other symptoms will be considered for predictors.

## Load packages
```{r}
#| output: false
library(tidyverse)
library(ggplot2)
library(here)
library(gtsummary)
```


## Load data
We will load the cleaned data from the `data` folder. To make coding easier, let's rename `cleandata` as `flu`
```{r}
#load cleaned data
filelocation <- here("fluanalysis", "data", "cleandata.rds")
load(filelocation)

#rename dataset
flu <- cleandata
```

## Overview of data
First, we can look at the proportion of responses that said they had the symptom using `tbl_summary()` function from the `gtsummary` package.
```{r}
#create summary table
sumtable <- tbl_summary(flu)

sumtable
```
Body temperature had a median temperature of 98.50 for all responses, and 35% of respondents said they has nausea. 
The most common symptoms were `cough`, `fatigue`,and `weakness` which was present in over 90% of responses. `CoughIntensity`, `Myalgia`, and `Weakness` all had multiple levels within the responses. 

We can see similar information using the `summary` function.
```{r}
summary(flu)
```
With this function, `BodyTemp` is further described to show a maximum of 103.10 degrees, but based on the other statistical information, the observations are mostly below 100 degrees.

## Plotting outcomes and potential predictors

We can start looking at the outcome variables using visuals and plots.

### Body temperature
Starting the body temperature, we will plot the density curve of `BodyTemp`. 
```{r}
#density curve of body temperature
ggplot(flu, aes(BodyTemp))+
  geom_density()
```
Similar to the statistical summary, the density curve shows that most data points are between 97.5 and 99.5 with the most observations occurring at ~ 98.6 degrees. 

### Body temperature and weakness
We will first consider the relationship between body temperature and weakness using the `geom_violin` function.
```{r}
#plot body temperature against weakness
ggplot(flu, aes(BodyTemp, Weakness, fill = Weakness))+
  geom_violin()
```
The "None" category appears to have the lowest average body temperature, "Mild" and "Moderate" appear to have similar distributions, and "Severe" category has the highest average temperature. Each category have observations with high temperatures, shown by the right side tails. 

### Body temperature and cough intensity

Next, body temperature and cough intensity will be viewed using the `geom_boxplot` function.
```{r}
#plot body temperature and cough intensity
ggplot(flu, aes(BodyTemp, CoughIntensity, fill = CoughIntensity))+
  geom_boxplot()
```
Similar to weakness, the average body temperature seems to be highest in the "Severe" category. However, the medians for all these categories are similar. 

### Body temperature and myalgia

Next, looking at body temperature and myalgia. 
```{r}
#plot body temperature and myalgia
ggplot(flu, aes(BodyTemp, Myalgia, fill = Myalgia))+
  geom_violin()
```
Average body temperatures are similar for each category compared to the other variables we have looked at. 

### Body temperature and runny nose

We can also consider a didactic variable (Present/Absent). Let's consider `RunnyNose`. 
```{r}
#plot body temperature and runny nose
ggplot(flu, aes(BodyTemp, RunnyNose, fill = RunnyNose))+
  geom_violin()
```
With this variable, there appears to be little difference between runny nose categories, but the "Yes" distribution was concentrated heavily around approx. 98.3 degrees. 

### Nausea

Let's move on to the categorical outcome variable of interest.
```{r}
#plot nausea by count
ggplot(flu, aes(Nausea, fill = Nausea))+
  geom_bar()
```
We can see about twice as many respondents reported no nausea. 

### Nausea and weakness

Considering nausea by level of weakness
```{r}
#plot nausea by weakness
ggplot(flu, aes(Nausea, fill= Weakness))+
  geom_bar(position = "dodge") # use position dodge to have bars side-by-side
```
Of those reporting nausea, mild weakness was the most common group. However, comparing the same level of weakness across the level of nausea shows that severe weakness had the highest proportion of nausea reported. 

### Nausea and cough intensity

Considering nausea and cough intensity:
```{r}
#plot nausea by cough intensity
ggplot(flu, aes(Nausea, fill= CoughIntensity))+
  geom_bar()+
  facet_wrap(.~CoughIntensity) #to compare levels of cough intensity
```
Considering respondents who reported nausea, the moderate weakness group reported nausea most often. Comparing within weakness levels, there appears to be a consistent 2:1 ratio of reporting no nausea to nausea present. 

### Nausea and myalgia

Considering nausea and myalgia
```{r}
#plot nausea by myalgia
ggplot(flu, aes(Nausea, fill= Myalgia))+
  geom_bar()+
  facet_wrap(.~Myalgia)
```
The group of interest here would be the severe myalgia group since there appears to be an even split of respondents reporting nausea. The other categories follow a similar trend as other leveled variables. 

### Nausea and runny nose

Considering nausea and runny nose (didactic variable):
```{r}
#plot nausea by runny nose
ggplot(flu, aes(Nausea, fill= RunnyNose))+
  geom_bar()+
  facet_wrap(.~RunnyNose)
```
Approximately half of the respondents in each group of runny nose (present or absent) reported nausea.

Now that we have explored the data, we can start model fitting. 






